Model-based Clustering for Neural Populations

### 2021-08-29 documents/cluster_neural_population.tex
document draft 1 complete

\section{Simulations}
There are two set of simulation examples. The first example is generated from LDS model directly, while in the second example the latents are generated directly without explicit specifying linear dynamics. In all of the following results, the covariance of process noise \(\mathbf{Q}\) is assumed to be block diagonal. The code for non-constrained and diagonal \(\mathbf{Q}\) can be found in XXX and XXX. The fitting results for all these three are similar, because the underlying true \(\mathbf{Q}\) is diagonal.

@@

\subsection{Simulation 1: Generate from LDS Directly}
In this simulation, there are 3 clusters with 10 neurons in each cluster. The dimension of latents in each cluster is 2. In the linear dynamics, the bias term \(\mathbf{b}\) is zero. There's no within-population interaction but has some weak between-population interactions. In other words, the linear dynamics matrix \(\mathbf{A}\) is roughly diagonal. The details of simulation can be found in the first section of XXX.

\subsubsection{Labeled Data: No Clustering}
The code for fitting is XXX, and some results are in Figure \ref{fig:LDS labeled}.

@@

\subsubsection{Unlabeled Data: Clustering}
To give the full path of clustering, I show results in GIFs. The code for FMM and DPMM can be found in XXX and XXX.

There are three results:
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Fit by FMM with \(\mathbf{J} = 3\), starting from random cluster assignments: 
	\item
	Fit by FMM with \(\mathbf{J} = 3\), starting from all neurons in single cluster: 
	\item
	Fit by DPMM (\(\alpha = X\)), starting from each neuron forms its own cluster:
\end{enumerate}
I didn't show the DPMM starting from single cluster. Since in my current implementation, the cluster generation is not efficient. In other words, the newly generated cluster will usually not be sampled and the algorithm usually gets stuck in 1 or 2 clusters.

@@

\subsection{Simulation 2: Generate Latents, without Specifying Linear Dynamics}
In this simulation, there are again 3 clusters with 2 latents in each. Besides set 10 neurons in each cluster, I further simulate 50 neurons in each cluster to see performance in a larger scale. Now, the latents are generated directly without specifying the underlying linear dynamics of latents. However, each latent is generated independently, so the linear dynamics matrix \(\mathbf{A}\) should be roughly diagonal. The details of simulation can be found in the first section of XXX.

\subsubsection{Labeled Data: No Clustering}
As in simulation 1, the data is fitted by the true cluster assignment (3 clusters) and forcing all neurons belong to single cluster. The code can be found in XXX.
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Smaller dataset (10 neurons each):\\
	The results with true cluster assignment is shown in Figure \ref{fig:no A labeled 10}
	\begin{figure}[h!]
		\makebox[\linewidth][c]{%
			\centering
			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{image006.png}}%
			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{image007.png}}%
			\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{image008.png}}%
		}
		\caption{unspecified dynamics with labels, 10 neurons each}
		\label{fig:no A labeled 10}
	\end{figure}
	And results by forcing all neurons to be in single cluster (Figure \ref{fig:no A labeled one cluster 10})
	\begin{figure}[h!]
		\makebox[\linewidth][c]{%
			\centering
			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.7\textwidth]{image009.png}}%
			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.7\textwidth]{image010.png}}%
		}
		\caption{unspecified dynamics with labels, one cluster, 10 neurons each}
		\label{fig:no A labeled one cluster 10}
	\end{figure}

@@

\subsubsection{Unlabeled Data: Clustering}
As previous, in each example, I show results from (1) FMM starting from random cluster assignment \(\left(J = 3\right) \), (2) FMM starting from single cluster and (3) DPMM starting from multiple clusters. The DPMM starting from single cluster need to be fixed later.

\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Smaller dataset (10 neurons each):\\
	\begin{enumerate}
		\def\labelenumi{\alph{enumi}.}
		\item
		Fit by FMM with \(\mathbf{J} = 3\), start from random cluster assignments:
		\item
		Fit by FMM with \(\mathbf{J} = 3\), start from single cluster:
		\item
		Fit by DP (\(\alpha = X\)), start from assuming each neuron forms its own cluster:
	\end{enumerate}

	\item
	Larger dataset (50 neurons each):\\
	\begin{enumerate}
		\def\labelenumi{\alph{enumi}.}
		\item
		Fit by FMM with \(\mathbf{J} = 3\), start from random cluster assignments:
		\item
		Fit by FMM with \(\mathbf{J} = 3\), start from single cluster:
		\item
		Fit by DP (\(\alpha = X\)), start from random assignment for 10 clusters:
	\end{enumerate}
\end{enumerate}

>>>>

\section{Simulations}
There are two set of simulation examples. The first example is generated from LDS model directly, while in the second example the latents are generated directly without explicit specifying linear dynamics. In all of the following results, the covariance of process noise \(\mathbf{Q}\) is assumed to be block diagonal. For reference, see the code for \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/old_example/sim_LDS_noConstraint.m}{non-constrained} and \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/old_example/sim_LDS_diag.m}{diagonal} \(\mathbf{Q}\). But these two legacy versions assuming the loading is independent on cluster assignments. If we need to use any of one of them, make sure to replace the updates for loading to be cluster-related. The fitting results for all these three are similar, because the underlying true \(\mathbf{Q}\) is diagonal.

@@

\subsection{Simulation 1: Generate from LDS Directly}
In this simulation, there are 3 clusters with 10 neurons in each cluster. The dimension of latents in each cluster is 2. In the linear dynamics, the bias term \(\mathbf{b}\) is zero. There's no within-population interaction but has some weak between-population interactions. In other words, the linear dynamics matrix \(\mathbf{A}\) is roughly diagonal. The details of simulation can be found in the first section of the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample.m}{simulation 1}.

\subsubsection{Labeled Data: No Clustering}
The code for fitting is \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample.m}{here}, and some results are in Figure \ref{fig:LDS labeled}.

@@

\subsubsection{Unlabeled Data: Clustering}
To give the full path of clustering, I show results in GIFs. Click to see the code for \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample_MM.m}{FMM} and \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/lds_sample_DP.m}{DPMM}.

There are three results:
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Fit by FMM with \(J = 3\), starting from random cluster assignments:
	\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/lds_samp_10_MM_above.gif}{GIF result 1}
	\item
	Fit by FMM with \(J = 3\), starting from all neurons in single cluster: 
	\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/lds_samp_10_MM_below.gif}{GIF result 2}
	\item
	Fit by DPMM (\(\alpha = 10\)), starting from each neuron forms its own cluster:
	\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/lds_samp_10_DP_above_10.gif}{GIF result 3}

\end{enumerate}
I didn't show the DPMM starting from single cluster. Since in my current implementation, the cluster generation is not efficient. In other words, the newly generated cluster will usually not be sampled and the algorithm usually gets stuck in 1 or 2 clusters.

@@

\subsection{Simulation 2: Generate Latents, without Specifying Linear Dynamics}
In this simulation, there are again 3 clusters with 2 latents in each. Besides set 10 neurons in each cluster, I further simulate 50 neurons in each cluster to see performance in a larger scale. Now, the latents are generated directly without specifying the underlying linear dynamics of latents. However, each latent is generated independently, so the linear dynamics matrix \(\mathbf{A}\) should be roughly diagonal. The details of simulation can be found in the \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample.m}{code}.

\subsubsection{Labeled Data: No Clustering}
As in simulation 1, the data is fitted by the true cluster assignment (3 clusters) and forcing all neurons belong to single cluster. The code can be found \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample.m}{here}.
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Smaller dataset (10 neurons each):\\
	The results with true cluster assignment is shown in Figure \ref{fig:no A labeled 10}
	\begin{figure}[h!]
		\makebox[\linewidth][c]{%
			\centering
			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{image006.png}}%
			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{image007.png}}%
			\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{image008.png}}%
		}
		\caption{unspecified dynamics with labels, 10 neurons each}
		\label{fig:no A labeled 10}
	\end{figure}
	And results by forcing all neurons to be in single cluster (Figure \ref{fig:no A labeled one cluster 10})
	\begin{figure}[h!]
		\makebox[\linewidth][c]{%
			\centering
			\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.7\textwidth]{image009.png}}%
			\subfigure[latents]{\label{fig:b}\includegraphics[width=0.7\textwidth]{image010.png}}%
		}
		\caption{unspecified dynamics with labels, one cluster, 10 neurons each}
		\label{fig:no A labeled one cluster 10}
	\end{figure}

@@

\subsubsection{Unlabeled Data: Clustering}
As previous, in each example, I show results from (1) FMM starting from random cluster assignment \(\left(J = 3\right) \), (2) FMM starting from single cluster and (3) DPMM starting from multiple clusters. The DPMM starting from single cluster need to be fixed later. These are code for
\href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample_MM.m}{FMM} and \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/unspecifiedA_sample_DP.m}{DPMM}.

\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Smaller dataset (10 neurons each):\\
	\begin{enumerate}
		\def\labelenumi{\alph{enumi}.}
		\item
		Fit by FMM with \(J = 3\), start from random cluster assignments:
		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_10_MM_above.gif}{GIF result 4}
		\item
		Fit by FMM with \(J = 3\), start from single cluster:
		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_10_MM_below.gif}{GIF result 5}
		\item
		Fit by DP (\(\alpha = 5\)), start from assuming each neuron forms its own cluster:
		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_10_DP_above_5.gif}{GIF result 6}
	\end{enumerate}

	\item
	Larger dataset (50 neurons each):\\
	\begin{enumerate}
		\def\labelenumi{\alph{enumi}.}
		\item
		Fit by FMM with \(\mathbf{J} = 3\), start from random cluster assignments:
		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_50_MM_above.gif}{GIF result 7}
		\item
		Fit by FMM with \(\mathbf{J} = 3\), start from single cluster:
		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_50_MM_below.gif}{GIF result 8}
		\item
		Fit by DP (\(\alpha = 1\)), start from random assignment for 10 clusters:
		\href{https://github.com/weigcdsb/state-space-clustering/blob/main/results/gif/noA_50_DP_above1.gif}{GIF result 9}
	\end{enumerate}
\end{enumerate}
